---
title: "Import data: CSV, TXT and Excel"
output: html_notebook
---

<!-- 
This file by Martin Monkman is licensed under a Creative Commons Attribution-ShareAlike 4.0 International (CC BY-SA 4.0). 
https://creativecommons.org/licenses/by-sa/4.0/
-->



```{r setup}
library(tidyverse)
library(readxl)
```




## 1. COVID-19 in BC

Use what you know about the functions in the {bcdata} package to import the details about COVID-19 cases in British Columbia. Assign the data to an R dataframe object with a meaningful name.


### solution

First search the data catalogue

```{r}
bcdc_search("COVID")
```

Then access the data.

```{r}
covid_cases <- bcdc_get_data("b-c-covid-19-case-details")
```



## REFERENCE MATERIAL

_R for Data Science_, "Data Import":
https://r4ds.had.co.nz/data-import.html


_R Cookbook (2nd ed.)_, "Input and Output":
https://rc2e.com/inputandoutput



### Package reference pages

{readr} 
https://readr.tidyverse.org/

{readxl}
https://readxl.tidyverse.org/index.html


