---
title: "BIDA 405 - week 5"
subtitle: "homework assignment #5: solutions"
author: "Martin Monkman"
date: "2020/10/04"
output: html_notebook
---

```{r setup}

library(tidyverse)

```

# A. The computational environment

Q1. Use the {annotater} package, which creates an RStudio Addin, to record the package versions that are loaded into your environment.

https://github.com/luisDVA/annotater

Q2. Use the `R.Version()` function in an R code chunk to record what version of R you have running.

```{r}
#
rversion <- R.Version()
rversion$version.string
# or
version
```



# B. Codebook 

For this segment, create a codebook using the {palmerpenguins} data. Consider the use of factor variable types and the creation of metadata about the entire dataset.

The reference page for the {codebook} package is https://rubenarslan.github.io/codebook/ 

```{r}

penguins <- readr::read_csv("penguins.csv")

```

```{r}
# solution
# specify species and island as factors
penguins %<>% 
  mutate(species = as.factor(species),
         island = as.factor(island))


```



-30-
