---
title: "BIDA 405 - week 5"
subtitle: "homework assignment #5: solutions"
author: "Martin Monkman"
date: "2020/10/04"
output: html_notebook
---

```{r setup}

library(tidyverse) # Easily Install and Load the 'Tidyverse', CRAN v1.3.0 

library(annotater) # Annotate Package Load Calls, [github::luisDVA/annotater] v0.1.3 
library(fs) # Cross-Platform File System Operations Based on 'libuv', CRAN v1.4.1 
library(here) # A Simpler Way to Find Your Files, CRAN v0.1 

```

# A. The computational environment

For this part of the assignment, you will create a new R Markdown file that will capture three elements of the computational environment used for this homework assignment. 

NOTE: you may need to install the packages listed above. 

Step 1: Create a new R Markdown file, and save it to your week 5 homework folder. Give the file a descriptive name.

Step 2: Copy the R setup chunk of code above, with the `library()` function for all of the packages.

Step 3: Use the {annotater} package, which creates an RStudio Addin, to record the package versions that are loaded into your environment.

The function reference for annotater is here: https://github.com/luisDVA/annotater

Step 3: Use the `R.Version()` function in an R code chunk to record what version of R you have running.

```{r}
#
rversion <- R.Version()
rversion$version.string
# or
version
```

Step 4: Use the function `fs::dir_tree` to create a tree diagram of your week 5 homework folder.


```{r}

fs::dir_tree()

```


### _Important note!_ 

**{annotater}**

The {annotater} package is not on CRAN. In order to install it, you need to 

* Step 1: the package {remotes} with the following code:

`install.packages("remotes")`

* Step 2: install {annotater} from GitHub

`remotes::install_github("luisDVA/annotater")`

The {annotater} reference page is https://github.com/luisDVA/annotater

***

# B. Codebook 

For this segment, create a codebook using the {palmerpenguins} data. Consider the use of factor variable types and the creation of metadata about the entire dataset.

The reference page for the {codebook} package is https://rubenarslan.github.io/codebook/ 

```{r}

penguins <- readr::read_csv("penguins.csv")

```

```{r}
# solution
# specify species and island as factors
penguins %<>% 
  mutate(species = as.factor(species),
         island = as.factor(island))


```


***

# C. README 

Step 1: Create a new Markdown file and call it README.md or README.Rmd (if you created an R Markdown file).

* Note: you can create a raw markdown file in a text editor like textpad or online at https://dillinger.io/

Step 2: Write a short README that describes a BIDA 405 course materials "project" folder.

* What is the course?

* What are some key details about the course?

* Is there a logic to the folder structure that you want to describe?

Don't forget:

* Brevity counts...this doesn't have to be long. Bullets are good, and you don't need full sentences.

* Linkify



-30-




-30-



