---
title: "Import data: CSV, TXT and Excel"
output: html_notebook
---

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. 
Some of it comes straight from Hadley Wickham & Garrett Grolemund's book _R for Data Science_, found online here: https://r4ds.had.co.nz/
It uses great swaths of content from the tidyverse reference material for {readr} and {readxl}, by RStudio. 
You can find that material here: 
https://readr.tidyverse.org/index.html 
https://readxl.tidyverse.org/index.html
-->


# Getting started

Run this code chunk:

```{r}
library(tidyverse)  # we will be using the functions of {readr}
library(readxl)
```



# 1. Reading a CSV file

CSV files are very commonly used for storing flat files. They don't have any formatting—just the number or text in the cell.

"CSV" stands for "Comma Separated Values", letting you know that a comma is used to separate one variable from the next.

The **R** package {readr} is designed to make importing these files simple.


(This example comes straight from the reference material for the {readr} package, found at https://readr.tidyverse.org/index.html)


Run this code chunk to create an object called `mtcars`, from a CSV file of the same name.

```{r}
# example
mtcars <- read_csv("mtcars.csv")

```

Adding the `cols()` specification allows us to alter what {readr} has decided for us. For example, we could set the `cyl` variable to be an integer.

```{r}
# example
read_csv("mtcars.csv", 
         col_types = 
           cols(cyl = col_integer())
)

```


## Your Turn 1.1

Insert an R chunk and rerun the example above, but with the `am` and `gear` variables also set to integer.

```{r}
# solution
read_csv("mtcars.csv",
         col_types = 
           cols(cyl = col_integer(),
                am = col_integer(),
                gear = col_integer())
)

```




The {readr} package allows a lot of control over how the file is read. Of particular utility are 

* `na = ""` -- specify which values you want to be turned into `NA`

* `skip = 0` -- specify how many rows to skip 

* `n_max = Inf` -- the maximum number of records to read


## Your Turn 1.2

Read the first 5 rows of the "mtcars.csv" file.

```{r}
# solution
read_csv("mtcars.csv", 
         n_max = 5)


```



# 2. Reading a fixed-width file

Fixed-width files are a hold-over from the days when storage was expensive and/or on punch cards. This meant that specific columns in the table (or card) were assigned to a particular variable, and precious space was not consumed with a delimiter. Compression methods have since meant that a CSV file with unfixed variable lengths are more common, but in some big data applications, fixed-width files can be much more efficient.

If you ever have to deal with a fixed-width file, you will (or should!) receive a companion file letting you know the locations of each variable in every row.


Star Wars characters: https://rdrr.io/cran/dplyr/man/starwars.html

```{r}

starwars <- dplyr::starwars


starwars %>% 
  filter(films == "A New Hope")

lapply(starwars$films, `[[`, 1)



```


```{r}
detect_string <- function(your_list, vector_strings){
  lapply(your_list, function(x) {
    if(TRUE %in% str_detect(x, paste(vector_strings, collapse = "|"))){
      TRUE
    } else {FALSE}
  })
}

detect_string(starwars$films, "A New Hope")   

newhope <- starwars %>% 
  filter(detect_string(starwars$films, "A New Hope") == TRUE) 

newhope <- newhope %>% 
  select(name, height, mass, homeworld, species) 

newhope

write_csv(newhope, "newhope.csv")

```



```{r}
iris

gdata::write.fwf(iris, file = "iris.txt", 
                 width = (c(
                   3,  # Sepal.Length
                   3,   # Sepal.Width
                   3,   # Petal.Length
                   3,   # Petal.Width
                   10   # Species
                   ))
                 )


```




STOP HERE ...


# 3. Reading an Excel file

If anything, Excel files are more common than CSV and other plain-text data files. They seem to multiply like coat hangers in the closet...

And as we see in the article by Karl Broman & Kara Woo, ["Data Organization in Spreadsheets"](https://www.tandfonline.com/doi/full/10.1080/00031305.2017.1375989), they can encourage some ways of storing information that makes it hard for us to analyze. Excel files can also contain a wide variety of data format types. {readxl} tries to figure out what's going on, but like {readr}, it allows you to override some of those automatic decisions.


```{r}
deaths <- read_excel("deaths.xlsx")
deaths
```

What do you notice about the "Date of birth" and "Date of death" columns?


## Your Turn 2.1

Read in the "deaths" file, but use the `range = ` option to specify when to start reading the file:


```{r}
# Example
deaths <- read_excel("deaths.xlsx", 
                     range = "A1:B3")

deaths

# solution
deaths <- read_excel(
  "deaths.xlsx",
  range = "A5:F15"
)

deaths

```


Use the `skip =` and `n_max =` options to achieve the same thing:

```{r}

# solution
deaths <- read_excel(
  readxl_example("deaths.xlsx"),
  skip = 4,
  n_max = 10
)

deaths


```


## Your Turn 2.2

Excel files often (almost always?) have multiple sheets in them. It's possible to specify which one you want to use with the `sheets = ""` option. You can also use the position number (if you happen to know it). 

Note that if you don't specify the sheet, {readxl} will default to the first one.

The `excel_sheets()` function will tell you the name of the sheets in an Excel file.

```{r}
datasets <- "datasets.xlsx"

excel_sheets(datasets)

# Example
read_excel(datasets, "iris")

```

Now, read in the "mtcars" sheet using the name of the sheet.


```{r}

# Solution
read_excel(datasets, "mtcars")

```

And finally, read in the "quakes" sheet, using the position.

```{r}
# Solution
read_excel(datasets, 4)

```




## REFERENCE MATERIAL

You can find more details in _R for Data Science_:
https://r4ds.had.co.nz/data-import.html


{readr} 
https://readr.tidyverse.org/

{readxl}
https://readxl.tidyverse.org/index.html


