---
title: 'Validation and cleaning'
output:
  html_document:
    df_print: paged
---

<!-- 
This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. 
-->


# Getting started

We've read in our raw data. The next step is to make sure that the data is what we expect, and what we need for our analysis. We need to _clean the data_. Which doesn't sound like a lot of fun. Didn't we want to be data scientists or business intelligence experts or academic scientists to uncover the insights hiding in the data? 

> The act of cleaning data imposes values/judgments/interpretations upon data intended to allow downstream analysis algorithms to function and give results. That’s exactly the same as doing data analysis. In fact, “cleaning” is just a spectrum of reusable data transformations on the path towards doing a full data analysis. —Randy Au, 2020-09-15 ^[["Data Cleaning IS Analysis, Not Grunt Work"](https://counting.substack.com/p/data-cleaning-is-analysis-not-grunt)]


So what are we doing when we say we are "cleaning the data"? And how can we confirm that it is "cleaned" in the way that we have defined?

## Some things that go wrong

Underlying the belief that the data is "dirty" and needs cleaning is the assumption that something has gone wrong. So...what can go wrong?


What are common dirty data problems?

* Column headers are values instead of variable names, or there are duplicate variable names.

* Multiple data types in a column. For example, some records are numbers and some are character strings—and this could be entirely correct, but if some number entry includes commas as thousands separators, it might be stored as a character.

* Multiple variables are stored in one column (examples include first name & last name, city & province or state).

* Misspellings (typing errors) ^[In this tweet, you can see 57 different ways of spelling "Philadelphia" in the data collected in a US Government loan form(https://twitter.com/dataeditor/status/1280278987797942272?s=20)]

* Value in wrong field (e.g. country entered into `city` field).

* Irregularities in unit of measurement (in an international survey or form, `salary` completed by respondents in the values of their local currency).

* Default values in the place of missing values.

> Dr Davis Lawrence, director of safety-literature database the SafeLit Foundation...tells me that 'in most US states the quality of police crash reports is at best poor for use as a research tool. ... Data-quality checks were rare and when quality was evaluated it was found wanting. For example, in Louisiana for most crashes in the 1980s most of the occupants were males who were born on January 1st, 1950. Almost all of the vehicles involved in crashes were the 1960 model year.' Except they weren't. These were just the default settings. [@Criado_Perez_2019, p.190]

* Contradiction (for example, two databases with the same person but the date of birth differs, perhaps due to non-ISO8601 entry: 07-08-79 and 08-07-79 both have the same digits but one could be mm-dd-yy and the other dd-mm-yy...we just don't know which is the correct one. Or is one a typo?)


See [@Zumel_Mount_2019, chapters 3 and 4]


The first challenge: how do we find these things?

The second challenge: what can and should we do about them?


## Identifying dirty data

There are some approaches to finding problems in the data that can be automated (done programmatically). These are sometimes called "data validation" methods; they can be applied when the data is collected (for example, in a web-based survey tool) or after the data has been collected.

The analyst can specify parameters prior to it being made available for the next steps. These can be expressed as rules or assumptions, and are based on knowledge of the subject. 

^[Mark van der Loo and Edwin de Jonge, [_Statistical Data Cleaning with Applications in R_, Chapter 6: Data Validation](https://learning-oreilly-com.ezproxy.library.uvic.ca/library/view/statistical-data-cleaning/9781118897157/?ar?orpq&email=oJbr4Ho8VLXz0zbFzjLK2g%3d%3d&tstamp=1600297427&id=39EE56C0062775092344911C1657C64468AFDE9B)]

* A field such as postal code has to be structured in a certain way (this is where your regex skills come in)

* A value might have a certain range 

* Some values, such as age, have to be positive

* There might be a known relationship between two or more variables


### The {validate} package

As with other data analysis topics, people in the R user community have confronted the problem of dirty data, and have written and shared functions to help the rest of us. The {validate} package was written and is supported by two of the methodologists at Statistics Netherlands.

The first function we will investigate is `check_that`, using the data in {palmerpenguins}

```{r setup}
library(tidyverse)

library(palmerpenguins)
library(validate)

penguins

```


Range: we know, being experts on all things penguin, that gentoo penguins are the third largest of the penguin species, and can weigh up to 8.5 kg. So we might want to see if we have any records with penguins with a body mass of more than 8500 grams. 

```{r}

check_penguins <- check_that(penguins, body_mass_g <= 8500)

summary(check_penguins)

```

What we see is that none of the penguins have a body mass of greater than 8500 grams, and there are 2 "NA" values in the data.


Relationship: 

While the heaviest gentoo penguins can be up to 8.5 kg, it is very unusual to find an Adélie or Chinstrap penguin that weighs more than 5 kg.

In our script, we use an `if()` statement that filters out the gentoo penguins, then checks the body mass.

```{r}

check_penguins <- check_that(penguins, 
                             if (species != "Gentoo") body_mass_g <= 5000)

summary(check_penguins)

```

In addition to telling us that there are 61 records where the value in the `body_mass_g` variable is greater than 5000 grams, it is also letting us know that there is 1 `NA` values.

Let's explore the data a bit more...if we filter our penguins by size, what do we see?

```{r}
penguins %>%
  filter(body_mass_g > 5000) %>% 
  distinct(species)

```

So this is confirms that all of the heavier penguins are gentoo, meeting our expectations.


Relationship: we know that some species have only been measured on one of the three islands. Gentoo penguins have only been sampled on Biscoe Island, and chinstrap penguins in our sample come only from Dream Island. Does the data reflect that?

In the first test, if it's not a gentoo it gets included in the `passes`--and if it is a gentoo, it has to be on Biscoe Island to get a pass.


```{r}

check_penguins <- check_that(penguins, 
                             g = if (species == "Gentoo") island == "Biscoe",
                             c = if  (species == "Chinstrap") island == "Dream")

summary(check_penguins)

```

Both tests have all of the cases pass, and none fail. 

We could have written code to generate a summary table to check this. In many cases, however, there will be too many cases to investigate.

```{r}
penguins %>% 
  group_by(species, island) %>% 
  tally
```

A single test could be written using the "OR" operator `|`.

```{r}

check_penguins <- check_that(penguins, 
                             if (species == "Gentoo") island == "Biscoe" &
                             if  (species == "Chinstrap") island == "Dream")

summary(check_penguins)



```


We can also test against a list:

```{r}
island_list <- c("Biscoe", "Dream", "Torgersen")

check_penguins <- check_that(penguins, if (species == "Adelie") island %in% island_list)

summary(check_penguins)


```

_What happens if we fail to include Dream Island in our list of islands where Adélie penguins are found?_



Year

```{r}

check_penguins <- check_that(penguins, year >= 2007 & year <= 2009)
summary(check_penguins)

```

What if we check for a single year?

```{r}
check_penguins <- check_that(penguins, year == 2009)
summary(check_penguins)

```

The output gives us the number of cases in that year in the `passes` the test, and the number of cases that are not in that year in the `fails`.



```{r}
reallybadpenguins <- read_rds(here::here("data", "reallybadpenguins.rds"))

check_penguins <- check_that(reallybadpenguins, year >= 2007 & year <= 2009)
summary(check_penguins)

```

```{r}
check_penguins <- check_that(reallybadpenguins, island %in% island_list)
summary(check_penguins)
```


## REFERENCE MATERIAL

You can find more details in _R for Data Science_:
https://r4ds.had.co.nz/data-import.html


{readr} 
https://readr.tidyverse.org/

{readxl}
https://readxl.tidyverse.org/index.html


